---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggoncoplot

<!-- badges: start -->

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental) [![CRAN status](https://www.r-pkg.org/badges/version/ggoncoplot)](https://CRAN.R-project.org/package=ggoncoplot)

<!-- badges: end -->

**ggoncoplot** creates interactive oncoplots from mutation level datasets

## Installation

You can install the development version of ggoncoplot like so:

``` r
remotes::install_github('selkamand/ggoncoplot')
```

## Usage


### Input

The input for ggoncoplot is a data.frame with 1 row per mutation in cohort and columns describing the following:

- Gene Symbol

- Sample Identifier 

- (optional) mutation type

- (optional) tooltip (character string: what we show on mouse hover over a particular mutation)


These columns can be in any order, and named anything. You define the mapping of your input dataset columns to the required features in the call to **ggoncoplot**


```{r example}
library(dplyr)
library(ggoncoplot)

# TCGA GBM dataset from TCGAmuations package
gbm_csv <- system.file(package='ggoncoplot', "testdata/GBM_tcgamutations_mc3_maf.csv.gz")
gbm_df <- read.csv(file = gbm_csv, header=TRUE)

gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    topn = 5, 
    interactive = FALSE
  )
```

## Tooltip Customisation


Lets add a custom tooltip that descrones the specific mutation


```{r}
gbm_df |> 
  mutate(tooltip = paste0(Reference_Allele, ">", Tumor_Seq_Allele2)) |>
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    col_tooltip = 'tooltip', # We'll specify a custom tooltip based on our new 'tooltip' column
    topn = 10, 
    interactive = FALSE
  )
```

Note tooltips are html, so if you want to insert a break, just paste in `<br>`.

Similarly, if you want to make text in the tooltip bold, try `"<b>text_to_bold<\b>"`
