---
title: "Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quick_start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggoncoplot)
library(dplyr, warn.conflicts = FALSE)
```

## Input Data

The input for ggoncoplot is a `data.frame` with 1 row per mutation. `data.frame` must contain columns describing the following:

- Gene Symbol

- Sample Identifier 

- (optional) mutation type

- (optional) tooltip (character string: what we show on mouse hover over a particular mutation)


These columns can be in any order, and named anything. You define the mapping of your input dataset columns to the required features in the call to **ggoncoplot**



## Minimal Example

```{r}

# TCGA GBM dataset from TCGAmuations package
gbm_csv <- system.file(package='ggoncoplot', "testdata/GBM_tcgamutations_mc3_maf.csv.gz")
gbm_df <- read.csv(file = gbm_csv, header=TRUE)

gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode'
  )
```

## Colour by mutation type

Colour by mutation by specifying `col_mutation_type`

```{r}

# TCGA GBM dataset from TCGAmuations package
gbm_csv <- system.file(package='ggoncoplot', "testdata/GBM_tcgamutations_mc3_maf.csv.gz")
gbm_df <- read.csv(file = gbm_csv, header=TRUE)

gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification'
  )
```


## Control which genes are shown

### Show top [n] Genes

Show the 4 most frequently mutated genes using `topn` argument

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    topn = 4
  )
```

### Exclude Specific Genes

Use the `genes_to_ignore` argument to filter out specific genes, such as TTN and MUC16.

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    topn = 10,
    genes_to_ignore = c("TTN", "MUC16")
  )
```

You can use packages like [somaticflags](https://github.com/CCICB/somaticflags) to get lists of genes you might want to filter out.

### Gene Subset

lets only show TP53 and TERT

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    genes_to_include = c('TP53', 'TERT'),
  )
```

## Control what samples are shown

The `show_all_samples` argument will add samples that don't have mutations in the selected genes to the plot.

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    show_all_samples = TRUE
  )
```


Note that if you supply a metadata table, by default samples lacking ANY mutations at all will still not be shown. You can include these samples by setting `metadata_require_mutations = FALSE` but this isn't recommended unless you're sure the sample truly has no mutations at all in the dataframe.


## Custom Tooltip

Use the `col_tooltip` argument to indicate which column of your input dataframe should be used as a custom tooltip.

```{r}
gbm_df |> 
  mutate(tooltip = paste0(Chromosome, ":", Start_Position, " ", Reference_Allele, ">", Tumor_Seq_Allele2)) |>
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    col_tooltip = 'tooltip' # We'll specify a custom tooltip based on our new 'tooltip' column
  )
```

Note tooltips are html, so if you want to insert a break, just paste in `<br>`.

Similarly, if you want to make text in the tooltip bold, try `"<b>text_to_bold<\b>"`.

Note that where a single sample has multiple mutations in a gene, are represented as one tile in oncoplot, tooltip for each mutation are shown (newline delimited).

## Add margin plots

### Gene Barplot

How many samples have mutations in each Gene (optionally coloured by mutation type)

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification',
    draw_gene_barplot = TRUE
  )
```

### Tumour Mutation Burden

How many mutations (total) are in each sample

**Not implemented yet: coming soon (https://github.com/selkamand/ggoncoplot/issues/30)**

## Static Plots

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    interactive = FALSE
  )
```


