---
title: "Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quick_start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggoncoplot)
library(dplyr, warn.conflicts = FALSE)

# Used in 'interaction with other packages' section
library(express) # For Creating Expression Plots
library(patchwork) # For combining oncoplot and expression plots
library(ggiraph) # For making combined plot interactive
```

## Input Data

The input for ggoncoplot is a `data.frame` with 1 row per mutation. `data.frame` must contain columns describing the following:

- Gene Symbol

- Sample Identifier 

- (optional) mutation type

- (optional) tooltip (character string: what we show on mouse hover over a particular mutation)


These columns can be in any order, and named anything. You define the mapping of your input dataset columns to the required features in the call to **ggoncoplot**



## Minimal Example

```{r}

# TCGA GBM dataset from TCGAmuations package
gbm_csv <- system.file(package='ggoncoplot', "testdata/GBM_tcgamutations_mc3_maf.csv.gz")
gbm_df <- read.csv(file = gbm_csv, header=TRUE)

gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode'
  )
```

## Colour by mutation type

Colour by mutation by specifying `col_mutation_type`

```{r}

# TCGA GBM dataset from TCGAmuations package
gbm_csv <- system.file(package='ggoncoplot', "testdata/GBM_tcgamutations_mc3_maf.csv.gz")
gbm_df <- read.csv(file = gbm_csv, header=TRUE)

gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification'
  )
```


## Control which genes are shown

### Show top [n] Genes

Show the 4 most frequently mutated genes using `topn` argument

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    topn = 4
  )
```

### Exclude Specific Genes

Use the `genes_to_ignore` argument to filter out specific genes, such as TTN and MUC16.

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    topn = 10,
    genes_to_ignore = c("TTN", "MUC16")
  )
```

You can use packages like [somaticflags](https://github.com/CCICB/somaticflags) to get lists of genes you might want to filter out.

### Gene Subset

lets only show TP53 and TERT

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    genes_to_include = c('TP53', 'TERT'),
  )
```

## Control what samples are shown

The `show_all_samples` argument will add samples that don't have mutations in the selected genes to the plot.

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    show_all_samples = TRUE
  )
```


Note that if you supply a metadata table, by default samples lacking ANY mutations at all will still not be shown. You can include these samples by setting `metadata_require_mutations = FALSE` but this isn't recommended unless you're sure the sample truly has no mutations at all in the dataframe.


## Custom Tooltip

Use the `col_tooltip` argument to indicate which column of your input dataframe should be used as a custom tooltip.

```{r}
gbm_df |> 
  mutate(tooltip = paste0(Chromosome, ":", Start_Position, " ", Reference_Allele, ">", Tumor_Seq_Allele2)) |>
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    col_tooltip = 'tooltip' # We'll specify a custom tooltip based on our new 'tooltip' column
  )
```

Note tooltips are html, so if you want to insert a break, just paste in `<br>`.

Similarly, if you want to make text in the tooltip bold, try `"<b>text_to_bold<\b>"`.

Note that where a single sample has multiple mutations in a gene, are represented as one tile in oncoplot, tooltip for each mutation are shown (newline delimited).

## Add Pathway Annotations

We can also add pathway information to the oncoplot by supplying a simple 2-column data.frame.

Currently the default order of pathways and genes in the plot are based on their order of appearnce in the pathway data.frame. Future versions of ggoncoplot will support data-based sorting.
Any genes missing from the oncoplot will be displayed under an 'Other' pathway at the very bottom of the plot.

```{r}

path_pathways <- system.file("testdata/GBM_tcgamutations_mc3.pathways.csv", package = "ggoncoplot")
pathways_df <- read.csv(path_pathways)

gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    pathway = pathways_df
  )
```


## Add margin plots

### Gene Barplot

How many samples have mutations in each Gene (optionally coloured by mutation type)

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification',
    draw_gene_barplot = TRUE
  )
```

### Tumour Mutation Burden

You can use set `draw_tmb_barplot = TRUE` to plot the total number of mutations (total mutational burden) in each sample. In most datasets, the presence of one hypermutator will makes it hard to see less extreme trends, and so by defualt mutational burden is plotted on a log10 scale. This can be changed by setting `log10_transform_tmb = FALSE`

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification',
    draw_tmb_barplot = TRUE, 
    #log10_transform_tmb = FALSE
  )
```


### Add both TMB and and Gene Barplots

Usually, we'll want to draw both margin plots (tmb + gene).

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification',
    draw_tmb_barplot = TRUE, 
    #log10_transform_tmb = FALSE
  )
```

### Add clinical annotations

```{r}
gbm_clinical_csv <- system.file(
 package = "ggoncoplot",
 "testdata/GBM_tcgamutations_mc3_clinical.csv"
)

gbm_clinical_df <- read.csv(file = gbm_clinical_csv, header = TRUE)

ggoncoplot(
 gbm_df,
 col_genes = "Hugo_Symbol",
 col_samples = "Tumor_Sample_Barcode",
 col_mutation_type = "Variant_Classification",
 metadata = gbm_clinical_df,
 cols_to_plot_metadata = c('gender', 'histological_type', 'prior_glioma', 'tumor_tissue_site'), 
 draw_tmb_barplot = TRUE, 
 draw_gene_barplot = TRUE, 
 show_all_samples = TRUE
)
```


## Customising Interactivity

### Copy on Click

By default, clicking a tile of the oncoplot will copy the sample ID to clipboard. This can be customised using the `copy` argument. For example you can choose to copy the tooltip or gene instead.

```{r}

gbm_df |>
  ggoncoplot(
   col_genes = "Hugo_Symbol",
   col_samples = "Tumor_Sample_Barcode",
   col_mutation_type = "Variant_Classification",
   copy = 'gene' # see ?ggoncoplot for other valid values
  )
```
## Customising the Look

Want more control over look of an oncoplot? ggoncoplot takes an `options` argument to help control all visual paramaters. See `?ggoncoplot_options` for a full list of paramaters.

```{r}
gbm_df |> 
  ggoncoplot(
    # Data
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    draw_tmb_barplot = TRUE, 
    draw_gene_barplot = TRUE,
    
    #pathway = pathways_df,
    metadata = gbm_clinical_df,
     cols_to_plot_metadata = c('gender', 'histological_type', 'prior_glioma', 'tumor_tissue_site'),
    
    # Customise Visual Options
    options = ggoncoplot_options(
      
      # Interactive Plot Options
      interactive_svg_width = 12,
      interactive_svg_height = 6,

      # Relative the ratio of marginal plot size to main tile plot (% of total plot height/width)
      plotsize_gene_rel_width = 40, # Genebar plot takes 50% of plot width
      plotsize_tmb_rel_height = 30,
      plotsize_metadata_rel_height = 15,
  
      # Axis Titles
      xlab_title = "Glioblastoma Samples",
      ylab_title = "Top 10 mutated genes",
  
      # Fontsizes
      fontsize_xlab = 40,
      fontsize_ylab = 40,
      fontsize_genes = 16,
      fontsize_samples = 12,
      fontsize_count = 14,
      fontsize_tmb_title = 14,
      fontsize_tmb_axis = 11,
      fontsize_pathway = 16,
  
      # Customise Tiles
      tile_height = 1,
      tile_width = 1,
      colour_backround = "white",
      colour_mutation_type_unspecified = "grey10",
  
      # Show different elements
      show_sample_ids = FALSE,
      show_ylab_title = FALSE,
      show_xlab_title = FALSE,
      show_ylab_title_tmb = FALSE,
      show_axis_gene = TRUE,
      show_axis_tmb = TRUE,
  
      # Transformation and label scales
      log10_transform_tmb = TRUE,
      scientific_tmb = FALSE,
  
      # Pathway Faceting Colours / Text
      colour_pathway_text = "black",
      colour_pathway_bg = "white",
      colour_pathway_outline = "black",
      pathway_text_angle = 0,
  
      # Legend number of columns
      ggoncoplot_guide_ncol = 1
    )
)
```


## Static Plots

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    interactive = FALSE
  )
```



## Interaction with other packages

ggoncoplot outputs can be combined with other packages. Simply set interactive = FALSE so ggoncoplot returns a ggplot object instead of a ggiraph. Do the same for express. Then you can con

```{r}
## Create a Glioblastoma Oncoplot
ggoncoplot <- gbm_df |>
  ggoncoplot(
    col_genes = 'Hugo_Symbol',
    col_samples = 'Tumor_Sample_Barcode',
    col_mutation_type = 'Variant_Classification', 
    interactive = FALSE
  )


## Create a t-SNE describing the same GBM cohort
tsne <- express_precomputed_tsne("GBM", interactive = FALSE)

## Combine plots with patchwork
combined_plots <- wrap_plots(ggoncoplot, tsne, ncol = 1)

# View the interactive version with ggiraph
interactive_multiplot <- girafe(ggobj = combined_plots, height_svg = 6, width_svg = 10)

# Add some settings to choose how to make combined plots interactive
interactive_multiplot <- girafe_options(x = interactive_multiplot,
    opts_selection(type = "multiple", only_shiny = FALSE)
    )

interactive_multiplot
```


### Why does this work?

ggoncoplot when interactive = FALSE returns a ggplot object with the aesthetics required for ggiraph interactivity baked in (in this case, e.g. the 'data_id' is defined by values of col_sample). So long as other packages also produce a data_id aesthetic with matched values (using ggiraph package) + optionally tooltip & onclick aesthetics, then we can use patchwork to combine the plots thae


