---
title: "quick_start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quick_start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggoncoplot)
library(dplyr, warn.conflicts = FALSE)
```


## Input

The input for ggoncoplot is a data.frame with 1 row per mutation in cohort and columns describing the following:

- Gene Symbol

- Sample Identifier 

- (optional) mutation type

- (optional) tooltip (character string: what we show on mouse hover over a particular mutation)


These columns can be in any order, and named anything. You define the mapping of your input dataset columns to the required features in the call to **ggoncoplot**



## Minimal Example

```{r}

# TCGA GBM dataset from TCGAmuations package
gbm_csv <- system.file(package='ggoncoplot', "testdata/GBM_tcgamutations_mc3_maf.csv.gz")
gbm_df <- read.csv(file = gbm_csv, header=TRUE)

gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    #col_mutation_type = 'Variant_Classification', 
    # topn = 10, 
    # interactive = TRUE
  )
```

## Colour by mutation type

Colour by mutation by specifying `col_mutation_type`

```{r}

# TCGA GBM dataset from TCGAmuations package
gbm_csv <- system.file(package='ggoncoplot', "testdata/GBM_tcgamutations_mc3_maf.csv.gz")
gbm_df <- read.csv(file = gbm_csv, header=TRUE)

gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification'
  )
```


## Control which genes are shown

### Show TopN Genes

Show the 4 most frequently mutated genes using `topn` argument

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    topn = 4
  )
```

### Gene Subset

lets only show TP53 and TERT

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    genes_to_include = c('TP53', 'TERT'),
  )
```

## Custom Tooltip

Lets add a custom tooltip that describes the specific mutation

```{r}
gbm_df |> 
  mutate(tooltip = paste0(Reference_Allele, ">", Tumor_Seq_Allele2)) |>
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    col_tooltip = 'tooltip', # We'll specify a custom tooltip based on our new 'tooltip' column
    interactive = FALSE
  )
```

Note tooltips are html, so if you want to insert a break, just paste in `<br>`.

Similarly, if you want to make text in the tooltip bold, try `"<b>text_to_bold<\b>"`


## Static Plots

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    interactive = FALSE
  )
```


