---
title: "Manual"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
library(ggoncoplot)
```

**For the best viewing experience please access this manual from https://selkamand.github.io/ggoncoplot/articles/manual.html.**

## Installation

See [Installation & Quick Start](https://selkamand.github.io/ggoncoplot/index.html).


## Input data

The input for ggoncoplot is a **data.frame**/**data.table**/**tibble** with 1 row per mutation in
cohort and columns describing the following:

- Gene Symbol

- Sample Identifier 

- (optional) mutation type

- (optional) tooltip (character string: what we show on mouse hover over a particular mutation)


These columns can be in any order, and named anything. You define the mapping of your input dataset columns to the required features in the call to **ggoncoplot**.

### Finding public datasets

The [tidyTCGA](https://github.com/selkamand/tidyTCGA) package provides public tabular cancer datasets. ggoncoplot is flexible with input data; you can use MAF files or any other tabular mutation-level datasets by specifying the columns for sample identifiers and gene names.



## Minimal example

```{r}

# TCGA GBM dataset from TCGAmuations package
gbm_csv <- system.file(package='ggoncoplot', "testdata/GBM_tcgamutations_mc3_maf.csv.gz")
gbm_df <- read.csv(file = gbm_csv, header=TRUE)

gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode'
  )
```



## Colour by mutation type

Colour by mutation by specifying `col_mutation_type`

```{r}

# TCGA GBM dataset from TCGAmuations package
gbm_csv <- system.file(package='ggoncoplot', "testdata/GBM_tcgamutations_mc3_maf.csv.gz")
gbm_df <- read.csv(file = gbm_csv, header=TRUE)

gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification'
  )
```


## Control which genes are shown

### Show top [n] genes

Show the 4 most frequently mutated genes using `topn` argument

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    topn = 4
  )
```

### Exclude specific genes

Use the `genes_to_ignore` argument to filter out specific genes, such as TTN and MUC16.

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    topn = 10,
    genes_to_ignore = c("TTN", "MUC16")
  )
```

You can use packages like [somaticflags](https://github.com/CCICB/somaticflags) to get lists of genes you might want to filter out.

### Gene subset

lets only show TP53 and TERT

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    genes_to_include = c('TP53', 'TERT'),
  )
```

## Control what samples are shown

The `show_all_samples` argument will add samples that don't have mutations in the selected genes to the plot.

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    show_all_samples = TRUE
  )
```


Note that if you supply a metadata table, by default samples lacking ANY mutations at all will still not be shown. You can include these samples by setting `metadata_require_mutations = FALSE` but this isn't recommended unless you're sure the sample truly has no mutations at all in the dataframe.


## Customise tooltips

Use the `col_tooltip` argument to indicate which column of your input dataframe should be used as a custom tooltip.

```{r}

# Add a tooltip column to data.frame
gbm_df[["tooltip"]] <- with(gbm_df, paste0(Chromosome, ":", Start_Position, " ", Reference_Allele, ">", Tumor_Seq_Allele2))

# Generate oncoplot
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    col_tooltip = 'tooltip' # We'll specify a custom tooltip based on our new 'tooltip' column
  )
```

Note tooltips are html, so if you want to insert a break, just paste in `<br>`.

Similarly, if you want to make text in the tooltip bold, try `"<b>text_to_bold<\b>"`.

Note that where a single sample has multiple mutations in a gene, are represented as one tile in oncoplot, tooltip for each mutation are shown (newline delimited).

## Add pathway annotations

We can also add pathway information to the oncoplot by supplying a simple 2-column data.frame.

Currently the default order of pathways and genes in the plot are based on their order of appearnce in the pathway data.frame. Future versions of ggoncoplot will support data-based sorting.
Any genes missing from the oncoplot will be displayed under an 'Other' pathway at the very bottom of the plot.

```{r}
path_pathways <- system.file("testdata/GBM_tcgamutations_mc3.pathways.csv", package = "ggoncoplot")
pathways_df <- read.csv(path_pathways)

gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    pathway = pathways_df
  )
```


## Draw marginal plots (Gene counts +  TMB)

### Gene barplot

How many samples have mutations in each Gene (optionally coloured by mutation type)

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    draw_gene_barplot = TRUE
  )
```



### Tumour mutation burden (TMB)

You can use set `draw_tmb_barplot = TRUE` to plot the total number of mutations (total mutational burden) in each sample. In most datasets, the presence of one hypermutator will makes it hard to see less extreme trends, and so by defualt mutational burden is plotted on a log10 scale. This can be changed by setting `log10_transform_tmb = FALSE`

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification',
    draw_tmb_barplot = TRUE, 
    #log10_transform_tmb = FALSE
  )
```


### Add both TMB and Gene Barplots

Usually, we'll want to draw both margin plots (tmb + gene).

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification',
    draw_tmb_barplot = TRUE, 
    draw_gene = TRUE
    # log10_transform_tmb = FALSE
  )
```

### Add clinical annotations

```{r}
gbm_clinical_csv <- system.file(
 package = "ggoncoplot",
 "testdata/GBM_tcgamutations_mc3_clinical.csv"
)

gbm_clinical_df <- read.csv(file = gbm_clinical_csv, header = TRUE)

ggoncoplot(
 gbm_df,
 col_genes = "Hugo_Symbol",
 col_samples = "Tumor_Sample_Barcode",
 col_mutation_type = "Variant_Classification",
 metadata = gbm_clinical_df,
 cols_to_plot_metadata = c('gender', 'histological_type', 'prior_glioma', 'tumor_tissue_site'), 
 draw_tmb_barplot = TRUE, 
 draw_gene_barplot = TRUE, 
 show_all_samples = TRUE
)
```


### Change position of clinical annotations

Sometimes you want your clinical annotations above your oncoplot. This can be achieved by setting `metadata_position = "top"` 

```{r}
ggoncoplot(
 gbm_df,
 col_genes = "Hugo_Symbol",
 col_samples = "Tumor_Sample_Barcode",
 col_mutation_type = "Variant_Classification",
 metadata = gbm_clinical_df,
 cols_to_plot_metadata = c('gender', 'histological_type', 'prior_glioma', 'tumor_tissue_site'), 
 draw_tmb_barplot = FALSE, 
 draw_gene_barplot = TRUE, 
 show_all_samples = TRUE,
 options = ggoncoplot_options(metadata_position = "top")
)
```

### Sorting by clinical annotations

To sort an oncoplot clinical annotations specify the columns of your metadata data.frame that you want to sort by using the `metadata_sort_cols` argument. Specifying multiple columns (as shown below) will do a hierarchical/stratified sort (sorts on first column, then by second column, etc).


```{r}
ggoncoplot(
 gbm_df,
 col_genes = "Hugo_Symbol",
 col_samples = "Tumor_Sample_Barcode",
 col_mutation_type = "Variant_Classification",
 metadata = gbm_clinical_df,
 cols_to_plot_metadata = c('gender', 'histological_type', 'prior_glioma', 'tumor_tissue_site'), 
 draw_tmb_barplot = TRUE, 
 draw_gene_barplot = TRUE, 
 show_all_samples = TRUE,
 metadata_sort_cols = c("gender", "histological_type") # Sort by gender, then histological type
)
```

For finer control over how the sort is performed, see `metadata_sort_desc` and `metadata_sort_by` arguments. 

## Customising interactivity

### Copy on click

By default, clicking a tile of the oncoplot will copy the sample ID to clipboard. This can be customised using the `copy` argument. For example you can choose to copy the tooltip or gene instead.

```{r}
gbm_df |>
  ggoncoplot(
   col_genes = "Hugo_Symbol",
   col_samples = "Tumor_Sample_Barcode",
   col_mutation_type = "Variant_Classification",
   copy = 'gene' # see ?ggoncoplot for other valid values
  )
```

## Customising the look

### Inbuilt options (recommended)

Want more control over look of an oncoplot? ggoncoplot takes an `options` argument to help control all visual parameters. See `?ggoncoplot_options` for a full list of parameters.

```{r}
gbm_df |> 
  ggoncoplot(
    # Data
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    draw_tmb_barplot = TRUE, 
    draw_gene_barplot = TRUE,
    
    #pathway = pathways_df,
    metadata = gbm_clinical_df,
     cols_to_plot_metadata = c('gender', 'histological_type', 'prior_glioma', 'tumor_tissue_site'),
    
    # Customise Visual Options
    options = ggoncoplot_options(
      
      # Interactive Plot Options
      interactive_svg_width = 12,
      interactive_svg_height = 6,

      # Relative the ratio of marginal plot size to main tile plot (% of total plot height/width)
      plotsize_gene_rel_width = 40, # Genebar plot takes 50% of plot width
      plotsize_tmb_rel_height = 30,
      plotsize_metadata_rel_height = 15,
  
      # Axis Titles
      xlab_title = "Glioblastoma Samples",
      ylab_title = "Top 10 mutated genes",
  
      # Fontsizes
      fontsize_xlab = 40,
      fontsize_ylab = 40,
      fontsize_genes = 16,
      fontsize_samples = 12,
      fontsize_count = 14,
      fontsize_tmb_title = 14,
      fontsize_tmb_axis = 11,
      fontsize_pathway = 16,
  
      # Customise Tiles
      tile_height = 1,
      tile_width = 1,
      colour_backround = "white",
      colour_mutation_type_unspecified = "grey10",
  
      # Show different elements
      show_sample_ids = FALSE,
      show_ylab_title = FALSE,
      show_xlab_title = FALSE,
      show_ylab_title_tmb = FALSE,
      show_axis_gene = TRUE,
      show_axis_tmb = TRUE,
  
      # Gene Barplot Specific Options
      show_genebar_labels = TRUE,
      genebar_label_padding = 0.1,
      genebar_only_pad_when_labels_shown = TRUE,
      genebar_label_nudge = 3,
      genebar_label_round = 0,

      # Transformation and label scales
      log10_transform_tmb = TRUE,
      scientific_tmb = FALSE,
  
      # Pathway Faceting Colours / Text
      colour_pathway_text = "black",
      colour_pathway_bg = "white",
      colour_pathway_outline = "black",
      pathway_text_angle = 0,
  
      # Legend number of columns
      ggoncoplot_guide_ncol = 1
    )
)
```

### Standard patchwork styling (static plots only)

When building an oncoplot with `interactive=FALSE` the result is a `patchwork` object. This lets you customise the plot with [standard patchwork styling](https://patchwork.data-imaginist.com/articles/guides/annotation.html#styling-the-patchwork). For example adding a title with `plot_annotation()` or applying a theme to all subplots (tilegraph & gene barplot) with `& theme(...)`.

```{r}
library(patchwork) 
library(ggplot2) 

# Create oncoplot patchwork
oncoplot_patchwork <- gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode',
    draw_gene_barplot = TRUE,
    interactive = FALSE
  )

# Customise the theme
oncoplot_patchwork +
  plot_annotation(title = 'My Glioblastoma Oncoplot') & 
  theme(text = element_text('mono', face = "bold"))
```

## Static plots

```{r}
gbm_df |> 
  ggoncoplot(
    col_genes = 'Hugo_Symbol', 
    col_samples = 'Tumor_Sample_Barcode', 
    col_mutation_type = 'Variant_Classification', 
    interactive = FALSE
  )
```



## Interaction with other packages

ggoncoplot outputs can be combined with other packages to create interactive multiplots. Here we show how to co-explore mutation oncoplots, expression t-SNE's & methylation UMAPs. 


### Install additional packages required to generate interactive multiplots

The following example uses some additional packages: 

* [express](https://github.com/selkamand/express) for creating expression plots
* [patchwork](https://github.com/thomasp85/patchwork) for combining oncoplot, expression and methylation plots into a single graphic
* [ggiraph](https://github.com/davidgohel/ggiraph) for making the combined plot interactive
* [somaticflags](https://github.com/CCICB/somaticflags) for excluding genes commonly mutated in somatic tissues from the oncoplot.

To install these packages run:

```{r eval=FALSE, include=FALSE}
install.packages("somaticflags")
remotes::install_github('selkamand/express')

## ggiraph & patchwork should have been automatically installed alongside ggoncoplot,
## but you can also install by running: 
install.packages(c("ggiraph","patchwork"))
```

### Generate an interactive multiplot

To create an oncoplot that can be interactively coexplored with other visualisations, set `interactive = FALSE` so ggoncoplot returns a ggplot object instead of a ggiraph. Do the same for the express package which visualises TCGA methylation and expression data (as UMAPs and t-SNEs. Then you can use patchwork to compose the plots into a single visualisation make the entire group interactive with data linked across plots.

```{r}

# Load required libraries
library(express) # For Creating Expression Plots
library(patchwork) # For combining oncoplot and expression plots
library(ggiraph) # For making combined plot interactive
library(somaticflags) # For excluding genes commonly mutated in somatic tissues

## Create a Breast Cancer Oncoplot
brca_df <- read.csv(system.file("testdata/BRCA_tcgamutations_mc3.csv.gz", package = "ggoncoplot"))

brca_clinical <- read.csv(system.file("testdata/BRCA_tcgamutations_mc3_clinical.csv.gz", package = "ggoncoplot"))

ggoncoplot <- brca_df |>
  ggoncoplot(
    col_genes = 'Gene',
    col_samples = 'Sample',
    col_mutation_type = 'MutationType',
    topn = 10, 
    genes_to_ignore = somaticflags,
    metadata = brca_clinical,
    metadata_palette =
      list(
        Progesterone = c("Indeterminate" = "gray80", "Negative" = "black", "Positive" = "#DF536B", "[Not Evaluated]" = "grey90"),
        Estrogen = c("Indeterminate" = "gray80", "Negative" = "black", "Positive" = "#DF536B", "[Not Evaluated]" = "grey90"),
        HER2 = c("Indeterminate" = "gray80", Equivocal =  "grey80", "Negative" = "black", "Positive" = "#DF536B", "[Not Evaluated]" = "grey90"),
        Classification = c("Ambiguous" = "gray80", "Triple Negative" = "black", "Not Triple Negative" = "#ff0000")
        ),
    options = ggoncoplot_options(
      show_genebar_labels = TRUE, 
      plotsize_metadata_rel_height = 40, 
      plotsize_tmb_rel_height = 10, 
      genebar_label_nudge = 20, 
      fontsize_genes = 11,
      fontsize_metadata_text = 11,
      show_legend = FALSE
    ),
    interactive = FALSE,
    verbose=FALSE
  )


## Create a gene expression t-SNE describing the same BRCA cohort
tsne_expression <- express_precomputed("BRCA", datatype = "expression", interactive = FALSE)

## Create a methylation UMAP describing the same BRCA cohort
umap_methylation <- express_precomputed("BRCA", datatype = "methylation", interactive = FALSE)

## Combine plots with patchwork
combined_plots <- (tsne_expression + umap_methylation) / ggoncoplot  + plot_layout(heights = c(3.5, 6.5))

# View the interactive version with ggiraph
interactive_multiplot <- girafe(ggobj = combined_plots, height_svg = 6, width_svg = 9)

# Add some settings to choose how to make combined plots interactive
interactive_multiplot <- girafe_options(x = interactive_multiplot,
    opts_selection(type = "multiple", only_shiny = FALSE, css = "opacity: 1"),
    opts_selection_inv(css = "opacity: 0.12")
    )

interactive_multiplot
```


### Why does this work?

ggoncoplot when interactive = FALSE returns a ggplot object with the aesthetics required for ggiraph interactivity baked in (in this case, e.g. the 'data_id' is defined by values of col_sample). So long as other packages also produce a data_id aesthetic with matched values (using ggiraph package) + optionally tooltip & onclick aesthetics, then we can use patchwork to combine the plots.

### Linking datapoints on custom plots to the oncoplot

You can link data between any custom ggplot and an oncoplot. Follow these steps:

1. Create your custom ggplot as usual.
2. Replace your geoms with interactive versions from ggiraph.
3. Add the `data_id` aesthetic, mapping it to the same sample identifiers used in ggoncoplot.
4. Combine the custom plot and the oncoplot using patchwork.
5. Pass the combined plot to the `girafe()` function from ggiraph.

This will ensure that the data points are linked between the plots.



## Saving your plot

In the current version of ggoncoplot, the download button for the interactive plot downloads a low-resolution image. We recommend the following alternatives for high-quality plots suitable for scientific publications:

- To save your non-interactive plot as a 300 dpi PNG, PDF, or SVG, use the `ggsaves` function from [ggsaves](https://github.com/selkamand/ggsaves).

- To save your interactive plot in HTML, SVG, or PDF formats (from which high-resolution PNGs can be derived), use the `ggisaves` function from [ggsaves](https://github.com/selkamand/ggsaves).

Alternatives:
- Export the non-interactive graph in any format using your preferred export method for R plots (e.g., RStudio GUI, or `ggplot2::ggsave()`).
- Export the interactive graph in any format using your preferred export method for R HTML widgets (e.g., RStudio GUI or `htmlwidgets::saveWidget()`).


